Verified and terminating dissections

### 2018-05-18 paper/main.tex
Changed todo formatting and fill-width.

% TODO notes
\usepackage[draft]{todonotes}

@@

\begin{abstract}
The functional programming paradigm encourages a style of programming based on
  the use of higher-order functions over inductively defined datatypes. A fold is
  the prototypical example of such a function. However, its use for computation
  comes at a price.  Folds for branching datatypes, such as binary trees, are by
  definition not tail recursive functions.

  McBride has proposed a method called
  \emph{dissection}\cite{McBride:2008:CLM:1328438.1328474}, to
  transform a fold into its tail-recursive counterpart. Nevertheless,
  it is not clear why the resulting function terminates, nor it is
  clear that the transformation preserves the fold's semantics. In
  this paper we fill the gap by providing a fully machine checked
  proof of the construction using the proof assistant Agda.
\end{abstract}

\include{ccs.tex}

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

@@

The proof fails because in \ensuremath{\Varid{aux}} both Zippers \ensuremath{\Varid{x}} and \ensuremath{\Varid{y}} might very well be
locations of leaves belonging to different trees as far we know. Thanks to the 
use of dependent types, the property that a Zipper represents a location inside a 
concrete tree can be made explicit at the type level.

@@

We write a relation that is enforced to only relate Zippers beloging to the same \ensuremath{\AD{Expr}} by
indexing the \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}}

@@

The concrete details of the relation follow very much the one we gave before
with the exception that every case has attached a new piece of information
specifying the concrete \ensuremath{\AD{Expr}} to which both Zippers stand as locations.

The indexed relation is now suitable for proving well foundedness because by
introducing the equalities we are able to show how the structure of the two
input values are related so it is straightforward to use recursion. In concrete,
the case we explained before now can be proven by learning that \ensuremath{(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})}
stands for a position on the left subtree, and \ensuremath{(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})} on the right  of a
common\ensuremath{\AI{Add}} node.

% \begin{code}
%  <-WF : forall (e : Expr) Well-founded ltOp
%  <-WF x = acc (aux x)
%     where
%       aux : ∀ (x : Zipper)
%           -> ∀ (y : Zipper) -> y < x -> Acc ltOp y
%       aux dotted(t2 , Left t1' :: s2) dotted(t1 , Right n t2' eq :: s1) (<-Right-Left eq1 eq2) = {!!}
%       aux ...
% \end{code}

@@

\subsection{Termination}
\subsection{Correctness}
\section{Regular universe}
  + Universe interpretation generic programming
  + Fixpoint
  + Example??
\section{Dissection}
  + Dissection in agda
  + Plug
  + Zipper up Zipper down
  + Make clear the separation between recursion in the functor level and
  the fix level
  + relation on dissection?

@@

>>>>

% TODO notes
%\usepackage[draft]{todonotes}
\usepackage{xcolor}
\newcommand\todo[1]{\textcolor{red}{\textbf{TODO:~}#1}}

@@

\begin{abstract}
% The functional programming paradigm encourages a style of programming based on
%   the use of higher-order functions over inductively defined datatypes. A fold is
%   the prototypical example of such a function. However, its use for computation
%   comes at a price.  Folds for branching datatypes, such as binary trees, are by
%   definition not tail recursive functions.

%   McBride has proposed a method called
%   \emph{dissection}\cite{McBride:2008:CLM:1328438.1328474}, to
%   transform a fold into its tail-recursive counterpart. Nevertheless,
%   it is not clear why the resulting function terminates, nor it is
%   clear that the transformation preserves the fold's semantics. In
%   this paper we fill the gap by providing a fully machine checked
%   proof of the construction using the proof assistant Agda.
  \todo{Write abstract}
\end{abstract}

\include{ccs}

%% Keywords
%% comma separated list
\keywords{datatype generic programming, catamorphisms, dissection,
  dependent types, Agda, \todo{Keywords?}}

@@

The proof fails because in \ensuremath{\Varid{aux}} both Zippers \ensuremath{\Varid{x}} and \ensuremath{\Varid{y}} might very well be
locations of leaves belonging to different trees as far we know. Thanks to the
use of dependent types, the property that a Zipper represents a position inside a
concrete tree can be made explicit at the type level.

@@

We write a relation that is enforced to only relate Zippers beloging to the same
\ensuremath{\AD{Expr}} by using a common value of that type as an of \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}}

@@

The concrete details of the relation follow very much the one we gave before,
with the exception that every case has attached a new piece of information
specifying the concrete \ensuremath{\AD{Expr}} obtained by plugging both Zippers.

The new version of the relation is suitable for proving well foundedness because
we can pattern match on the equality included in the \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}} type to show how 
the overall structure decreases. This allows us to use the recursion we need to
complete the proof.
In particular, the case we were not able to prove before, now can be proven by learning that
\ensuremath{(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})} is a position on the left subtree while \ensuremath{(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})} is on the
right subtree of a common \ensuremath{\AI{Add}} node.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{<-WF}\;\mathbin{:}\;\forall\;\AB{e}\;\to \;\AF{Well\text{-}founded}\;(\Conid{IxltOp}\;\AB{e}){}\<[E]%
\\
\>[3]{}\Varid{<-WF}\;\Varid{x}\;\mathrel{=}\;\AI{acc}\;(\Varid{aux}\;\AB{e}\;\Varid{x}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AK{where}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{aux}\;\mathbin{:}\;\forall\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\to \;\forall\;(\Varid{x}\;\mathbin{:}\;\AD{Zipper\ensuremath{\Downarrow}}\;\AB{e}){}\<[E]%
\\
\>[7]{}\hsindent{4}{}\<[11]%
\>[11]{}\to \;\forall\;(\Varid{y}\;\mathbin{:}\;\AD{Zipper\ensuremath{\Downarrow}}\;\AB{e})\;\to \;\Varid{y}\;\AD{<}\;\Varid{x}\;\to \;\AD{Acc}\;(\Conid{IxltOp}\;\AB{e})\;\Varid{y}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{aux}\;\AS{.}\hspace*{-0.1cm}\;(\AI{Add}\;(\Varid{plug}))\;\AS{.}\hspace*{-0.1cm}\;((\AB{\ensuremath{t_2}}\;\Varid{,}\;\AI{Left}\;\AB{\ensuremath{t_1}'}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_2}}))\;\Varid{,}\;\AI{refl})\;\AS{.}\hspace*{-0.1cm}\;((\AB{\ensuremath{t_1}}\;\Varid{,}\;\AI{Right}\;\AB{n}\;\AB{\ensuremath{t_2}'}\;\AB{\ensuremath{eq}}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_1}})\;\Varid{,}\;\Varid{eq2})\;{}\<[E]%
\\
\>[7]{}\hsindent{4}{}\<[11]%
\>[11]{}(\AI{<\text{-}Right\text{-}Left}\;\Varid{eq1}\;\Varid{eq2})\;\mathrel{=}\;\{\mskip1.5mu \mathbin{!!}\mskip1.5mu\}\;{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{aux}\;\Varid{...}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

We have now the proof of well foundedness for the relation defined over top-down Zippers. We also have proven that
there is an equivalence between top-down and bottom-up Zippers. We exploit it by using the top-down encoding for
the termination proof while we use the bottom-up to actually compute in a tail recursive manner.

Thus we prove a lemma stating that if we apply unload to a bottom-up Zipper and this results in another
Zipper, then the result is smaller by the relation than the input. However, to show it we have to 
convert them to the top down representation. In overall, what we have is the following lemma:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{unload-ltop}\;\mathbin{:}\;\forall\;\AB{n}\;\AB{\ensuremath{eq}}\;\Varid{s}\;\Varid{t'}\;\Varid{s'}\;\to \;\AF{unload}\;(\Conid{Tip}\;\AB{n})\;(\Conid{TipA}\;\AB{n})\;\AB{\ensuremath{eq}}\;\Varid{s}\;\AD{\ensuremath{\equiv}}\;\Varid{inj₁}\;(\Varid{t'}\;\Varid{,}\;\Varid{s'}){}\<[E]%
\\
\>[B]{}\hsindent{13}{}\<[13]%
\>[13]{}\to \;(\Varid{t'}\;\Varid{,}\;\AF{reverse}\;\Varid{s'})\;\AD{\_<\_}\;(\AB{n}\;\Varid{,}\;\AF{reverse}\;\Varid{s}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\subsection{Correctness}

Indexing the \emph{Zipper} with an expression allow us to prove correcness of
the transformation easily. The expression during the fold does not change, thus
in every step of the computation the result of its evaluation remains constant.

By using induction over the definition of unload, we can prove that when \ensuremath{\AF{unload}} 
delivers a value, it corresponds to the result of evaluating of the input expression.
In order to do so, we enrich the type of \ensuremath{\AF{unload}} to include the expression that has already
been folded and we have its result.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{unload}\;\mathbin{:}\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\to \;(\AB{n}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}})\;\to \;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\AB{n}\;\to \;\AD{Stack}\;\to \;(\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack})\;\AD{\ensuremath{\uplus}}\;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\[\blanklineskip]%
\>[3]{}\Varid{unload-correct}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\forall\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;(\AB{n}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}})\;(\AB{\ensuremath{eq}}\;\mathbin{:}\;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\AB{n})\;(\Varid{s}\;\mathbin{:}\;\AD{Stack})\;(\Varid{x}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}}){}\<[E]%
\\
\>[19]{}\to \;\AF{unload}\;\AB{e}\;\AB{n}\;\AB{\ensuremath{eq}}\;\Varid{s}\;\Varid{≡}\;\AI{inj\ensuremath{_2}}\;\Varid{x}\;\to \;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Proving correctness of the whole transformation amounts to show that it holds for the
auxiliary recursor that we use to write the function \ensuremath{\AF{tail\text{-}rec\text{-}eval}}. We use well founded 
recursion to do structural recursion over the accesibility predicate and use the lemma 
\ensuremath{\Varid{unload-correct}} in the base case.

@@

\section{Regular universe}
  + Universe interpretation generic programming
  + Fixpoint
  + Example??
\section{Dissection}
  + Dissection in agda
  + Plug
  + Zipper up Zipper down
  + Make clear the separation between recursion in the functor level and
  the fix level
  + relation on dissection?

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% TeX-command-default: "lagda2pdf"
%%% End:
